---
title: 'Plotting with R: ggplot2'
author: "Yu-Zhen Tu <br> Research Assistant at NTU, GIBMS, Brain and Mind Laboratory <br> Community Mentor of JHU \"R Programming\" on Coursera <br> <a href=\"mailto:corytu@ntu.edu.tw\">corytu@ntu.edu.tw</a>"
date: "March 17, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", fig.align = "center")
```

# General Concepts

## Plotting systems in R (incompatible)
* base
    + Easy and straightforward (in my opinion)
    + Start with plot (or similar) functions
    + Use annotation functions to add elements or modify
    + Can't go back once the plot is created
* lattice
    + Useful for conditioning types of plots
    + Object oriented
    + Plots are created with one single function call
    + Intense preparation
* __ggplot2__
    + Combine the advatages of both

## Devices in R
* Screen
* Vector format files (e.g. pdf, svg)
* Bitmap format files (e.g. png, jpeg, tiff)

## Grammar of Graphics
* Mapping (creating a blank "map")
* Aesthetic (e.g. color, size)
* Geometric (e.g. points, lines, bars)

## Check R and ggplot2 versions
```{r, eval = FALSE}
version
# "version.string" >= 3.1.0
packageVersion("ggplot2")
# >= 2.2.0
```
* If your R version < 3.1.0, you may update your R on https://cloud.r-project.org.
* If your ggplot2 version < 2.2.0 or doesn't exist, please update or install your ggplot2 with the instruction in the next slide.

## Install and load ggplot2
```{r, eval = FALSE}
install.packages("ggplot2", repos = "https://cloud.r-project.org")
# "repos" argument is optional
```{r}
library(ggplot2)
```

# Scatter Plot <br> (iris & mtcars datasets)

## The structure of iris
```{r}
str(iris)
```

***
```{r}
g <- ggplot(aes(x = Petal.Length, y = Petal.Width), data = iris)
g <- g + geom_point(aes(color = Species)) +
  geom_smooth(method = "lm", color = "black")
g + labs(list(x = "Length of Petal", y = "Width of Petal",
              title = "My First Scatter Plot with ggplot2"))
```

## Save the plot
```{r, eval = FALSE}
ggsave("my_first_plot.png", width = 7, height = 7, dpi = 900)
# The file will be saved in your working directory unless specified
# 'plot = last_plot()' by default
```

## The structure of mtcars
```{r}
str(mtcars)
```

## It's your turn!
```{r, echo = FALSE}
g <- ggplot(aes(x = hp, y = mpg), data = mtcars)
g <- g + geom_point(aes(color = as.factor(cyl))) +
  geom_smooth(method = "lm", color = "black")
g + labs(list(x = "Gross Horsepower", y = "Miles/(US) Gallon",
              color = "Number of Cylinders", title = "Yes I Can!"))
```

# Bar Plots <br> (airquality dataset)

## The structure of airquality
```{r}
str(airquality)
```

## Data preparation
```{r}
se_cal <- function(x) {sd(x, na.rm = TRUE)/sqrt(length(na.omit(x)))}
myair <- sapply(1:4,
                function(i){tapply(airquality[,i], airquality$Month,
                                   mean, na.rm = TRUE)})
colnames(myair) <- names(airquality)[1:4]
myair <- as.data.frame(as.table(myair))
se <- sapply(1:4,
             function(i){tapply(airquality[,i], airquality$Month,
                                se_cal)})
myair$SE <- as.data.frame(as.table(se))$Freq
names(myair) <- c("Month", "Variable", "Mean", "SE")
```

***
```{r}
g <- ggplot(aes(x = Month, y = Mean), data = myair)
g <- g + geom_col(aes(fill = Variable), position = "dodge")
# geom_col() = geom_bar(stat = "identity")
g + geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE, group = Variable),
                       position = position_dodge(0.9), width = 0.2)
```

## Data for practice
```{r}
myiris <- sapply(1:4,
                 function(i){tapply(iris[,i], iris$Species,
                                    mean, na.rm = TRUE)})
colnames(myiris) <- names(iris)[1:4]
myiris <- as.data.frame(as.table(myiris))
se <- sapply(1:4,
             function(i){tapply(iris[,i], iris$Species, se_cal)})
myiris$SE <- as.data.frame(as.table(se))$Freq
names(myiris) <- c("Species", "Variable", "Mean", "SE")
```

## It's your turn!
```{r, echo = FALSE}
g <- ggplot(aes(x = Species, y = Mean), data = myiris)
g <- g + geom_col(aes(fill = Variable), position = "dodge")
g <- g + geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE,
                           group = Variable),
                       position = position_dodge(0.9), width = 0.25)
g + labs(list(y = "Mean Value (centimeters)",
              title = "Mean Length and Width versus Species"))
```

# Creating Facets

## Revisit the first scatter plot <br> (theme_classic)
```{r, echo = FALSE}
g <- ggplot(aes(x = Petal.Length, y = Petal.Width), data = iris)
g <- g + geom_point(aes(color = Species)) +
  geom_smooth(method = "lm", color = "black")
g <- g + labs(list(x = "Length of Petal", y = "Width of Petal",
                   title = "My First Scatter Plot with ggplot2"))
g <- g + theme_classic()
g
```

## Faceting
```{r}
g + facet_grid(. ~ Species)
```

## It's your turn!
```{r, echo = FALSE}
# mycars <- mtcars
# mycars$am <- sapply(1:nrow(mycars),
#                     function(i){
#                       ifelse(mycars$am[i] == 0, "Automatic", "Manual")})
g <- ggplot(aes(x = hp, y = mpg),
            data = transform(mtcars, am = factor(am, labels = c("Automatic", "Manual"))))
g <- g + geom_point(aes(color = as.factor(cyl))) +
  geom_smooth(method = "lm", color = "black")
g <- g + labs(list(x = "Gross Horsepower", y = "Miles/(US) Gallon",
                   color = "Number of Cylinders", title = "Yes I Can!"))
g <- g + theme_classic()
g + facet_grid(am ~ .)
```

# Extra Challenge

## Data simulation
```{r}
ID <- rep(1:20, each = 20)
age <- rep(c("Young", "Old"), each = 200)
sex <- rep(rep(c("Male", "Female"), each = 100), times = 2)
task <- rep(rep(c("A", "B"), each = 10), times = 20)
difficulty <- rep(1:10, times = 40)
simul <- data.frame(ID = ID,
                    age = factor(age, levels = c("Young", "Old")),
                    sex = factor(sex, levels = c("Male", "Female")),
                    task = factor(task, level = c("A", "B")),
                    difficulty = difficulty)
set.seed(1022)
simul$accuracy <- (rnorm(400, 80, 5) - 3*simul$difficulty -
  10*(as.numeric(simul$age)-1) +
  2*simul$difficulty*(as.numeric(simul$task)-1))/100
```

## Come on challengers!
```{r, echo = FALSE}
g <- ggplot(aes(x = difficulty, y = accuracy), data = simul)
g <- g + geom_point(aes(color = age, shape = sex))
g <- g + geom_smooth(aes(color = age, linetype = sex),
                     method = "lm")
g <- g + facet_grid(. ~ task) + theme_classic()
g + labs(list(x = "Difficulty", y = "Accuracy",
              color = "Age", shape = "Sex", linetype = "Sex"))
```

## Resources
* R in general
    + <a href="http://datascienceandr.org">R語言翻轉教室</a>
    + <a href="https://www.coursera.org/specializations/jhu-data-science">Data Science Specialization by JHU on Coursera</a>
* ggplot2 and coloring
    + <a href="http://ggplot2.org">ggplot2</a>
    + <a href="http://colorbrewer2.org">ColorBrewer</a>
* Asking questions
    + PTT R_Language, PTT Statistics
    + <a href="http://stackoverflow.com">Stack Overflow</a>, <a href="http://stats.stackexchange.com">Cross Validated</a>

# Thanks for your attention! <br> Happy plotting!