---
title: "R Tutorial"
author: "Yu-Zhen Tu (涂玉臻)<br><a href=\"mailto:corytu@ntu.edu.tw\">corytu@ntu.edu.tw</a><br>Research Assistant at Brain and Mind Laboratory, NTU<br>Community Mentor of JHU “R programming” on Coursera"
date: "August 15, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "", echo = TRUE)
```

## Outline
- Overview of R
- Input and evaluation
- Atomic classes of objects
- Data types
- Help documents
- Working directory
- Reading and looking at data
- Extracting subsets
- Vectorized operations
- Control structures
- Writing functions

# Overview of R
R as a language and a free software

## R programming language
- R is a dialect of S.
- S is a language developed by John Chambers at Bell Labs in 1976.
- Version 3 of S was rewritten in C language in 1988.
- R was created by Ross Ihaka and Robert Gentleman at the University of Auckland, New Zealand in 1991.
- The R Core Group was formed in 1997.
- R version 1.0.0 was released in 2000.
- R version 3.4.1 was released on June 30, 2017.

## R is a free software
- One is granted:
    1. The freedom to run the program for any purpose.
    2. The freedom to study how the program works.
    3. The freedom to share copies of the program.
    4. The freedom to improve or modify the program, and release the improvements to the public.
- “We call this free software because the user is free.” ([The Free Software Foundation](https://www.fsf.org))

# Input and Evaluation

## Input
- Use `<-` as the assignment operator.
- `=` is also eligible but I recommend `<-`. [Here](https://corytu.github.io/CourseraRMentoring/use-equal-or-arrow.html) is my reason for that.
- `#` starts comments. Comments won’t be executed.

    ```{r, eval = FALSE}
    x <- 5
    name <- "Yu-Zhen"  # keying my name in
    # Nothing happens.
    ```

- Incomplete expressions

## Evaluation and printing
```{r}
x <- 5    # inputting the value of x
x         # auto-printing
print(x)  # explicit printing
```

# Atomic Classes
Character, numeric, integer, complex, and logical

## Five basic classes of objects
- Character: `"Yu-Zhen"`, `letters`, `"Hello world!"`
- Numeric: `pi`, `123`, `Inf`
- Integer: `10L`, `123L`
- Complex: `7 + 3i`
- Logical: `TRUE` or `FALSE`
- Use the function `class` to find out which class the object is.

## Missing values
- `NA` (Not Available) and `NaN` (Not a Number)
- An `NaN` is an `NA`, but the converse is false.
- `is.na` and `is.nan` are used to test if objects are `NA` and `NaN`, respectively.

# Data Types
Vector, list, matrix, factor, and data frame

## Vector
- Use the function `c` (or `vector`) to create a vector.
- An integer sequence from m to n can be created by `m:n`.
- Vectors accept only the same class of elements.
- Coercion occurs when different classes are input. (logical < integer < numeric < complex < character)

    ```{r}
    c(TRUE, 1 + 3i, "Yay!")
    ```

- Explicit coercion can be performed with `as.`* functions.

## List
- Use `list` to create a list.
- Lists are similar with vectors but accept different classes of elements.

    ```{r, eval = FALSE}
    my_list <- list(1:3, letters, TRUE)
    my_list2 <- list(first = 1:3, second = letters, third = TRUE)
    # What's the difference between my_list and my_list2?
    ```

## Practice-1
Create a list like this:

```{r, echo = FALSE}
list(one = letters[1:4], two = 76:80, three = c(TRUE, FALSE, TRUE))
```

## Matrix
- Matrices are like vectors with two dimensions (row and column).
- Use `matrix` to create a matrix (column-wise by default).

    ```{r}
    matrix(1:6, 2, 3)
    ```

- Bind columns or rows by `cbind` or `rbind` respectively.

    ```{r}
    rbind(1:4, 5:8)
    ```

## Factor
- `factor` creates factors from vectors.
- The “levels” argument can be specified to assign base level.

    ```{r}
    factor(c("high", "high", "low", "mid", "low"),
           levels = c("low", "mid", "high"))
    ```

- [Further reading](https://corytu.github.io/CourseraRMentoring/factor-levels-and-labels.html) about factor levels
- Factors are very useful for analyzing categorical variables.
- Pay extra attention when coercing factors into numbers.

## Data frame
- `data.frame` can create one data frame.
- Data frames look like matrices but allow different classes of elements across columns.
- `cbind` and `rbind` are also effective for data frames.

## Practice-2
Try to make the following data frame. Save the object as "practice.frame".<br>*Hint: It's like creating a list but with the same length of vectors.*

```{r, echo = FALSE}
data.frame(num = 1:3,
           age = factor(c("young", "old", "young"),
                        levels = c("young", "old")),
           RT = c(12, 50, 15),
           MMSEpass = c(TRUE, FALSE, TRUE))
```

# Help Documents

## Looking up functions
- `?` or `help` gets the help document of the function.
    - Some arguments have default values.
    - Arguments can be passed by matching name or matching position.
- `??` or `help.search` finds functions with the key word.
- Getting more help
    - Google
    - R_Language on PTT
    - [Stack Overflow](https://stackoverflow.com)

# Working Directory

## Retrieve and set working directory
- Files will be saved to and retrieved from the working directory if no path is specified.
- `getwd` retrieves current working directory path.
- `setwd` sets another specific path as working directory.

## Practice
Let’s create a folder and set it as the working directory for this tutorial! Download the following two files and put them in that directory.<br>*Please avoid Chinese characters and spaces in your path.*

- [r-tutorial-demo.csv](https://github.com/corytu/RLanguagePresentations/blob/master/docs/r-tutorial-demo.csv)
- [r-tutorial-practice.csv](https://github.com/corytu/RLanguagePresentations/blob/master/docs/r-tutorial-practice.csv)

# Data Preprocessing
`read.table`/`write.table`;<br>`load`/`save.image`;<br>`source`/`dump`

## Reading tables
- `read.table` or `read.csv` inputs a table into R as a data frame.
- The important arguments include:
    - file: the file path or `file.choose`
    - header: whether the first row should be regarded as names
    - sep: specified separator for the table (e.g. “,” for .csv files)

    ```{r}
    demo.table <- read.csv("r-tutorial-demo.csv")
    ```

- `write.table` or `write.csv` outputs data in a spreadsheet.    

## Reading a saved workspace
- `load` inputs previously saved workspace into R.
- Workspace contains objects we’ve assigned.
- `save.image` saves the whole current workspace as a .RData file.

    ```{r, eval = FALSE}
    save.image("tutorial.RData")  # Save the whole workspace
    q()                           # Quit the R session
    # After turning on a new R session and setting the working directory
    load("tutorial.RData")
    ```

## Reading written codes
- `source` inputs a written .R file into R.
- `dump` can be used to save .R files.

## Looking at the data - 1
```{r}
demo.table
```

## Looking at the data - 2
```{r}
head(demo.table)
# Try tail(demo.table) by yourself
```

## Looking at the data - 3
```{r}
summary(demo.table)
```

## Looking at the data - 4
```{r}
str(demo.table)
```

## Practice
Take a look at the data frame you created earlier.
```{r, echo = FALSE}
str(data.frame(num = 1:3,
           age = factor(c("young", "old", "young"),
                        levels = c("young", "old")),
           RT = c(12, 50, 15),
           MMSEpass = c(TRUE, FALSE, TRUE)))
```

# Extracting Subsets

## Subsetting operators - 1
- `[]` extracts elements by names, indexes, or logical values.
    - It can be used to select more than one element.
    - It usually returns an object as the same class as the original.
- `[[]]` extracts elements from lists and data frames.
    - It can extract only one element.
    - Returned objects are not necessarily the same class as the original.
- `$` extracts elements by name.

## Subsetting operators - 2
- By default, subsetting a single row or a single column from a matrix will return a vector, unless the argument "drop" is FALSE.

    ```{r}
    demo.table$height
    # What if I only need the 2nd, 4th, and 5th elements of the heights?
    ```

# Vectorized Operations

## Vectorized operations in R

```{r, eval = FALSE}
x <- 1:3
y <- 4:6
# x + y = ?

a <- 1:4
b <- 1:2
# a - b = ?

m <- matrix(6, 2, 2)
n <- matrix(1:4, 2, 2)
# m * n = ?
# m %*% n = ?
```

## Practice
1. Read the .csv file, "r-tutorial-practice.csv", into your R session.
2. Look at the input data frame. Watch its components, structure, etc.
3. Revise the units of height and weight. (Height: from cm to m; Weight: from g to kg)
4. Calculate the BMIs of all the 11 participants and assign them as a new column called BMI of your data frame.
5. Report the BMIs of those who have MMSE passed.
6. Which participants have BMI greater than 25?
7. Export the revised data frame as a new csv file called "my-success.csv" to your working directory.

# Control Structures
`if`, `for`, `while`. These loops can be nested.

## Relational and logical operators
- `==` (equal to)
- `>=` (greater than or equal to)
- `<=` (less than or equal to)
- `>` (greater than)
- `<` (less than)
- `!=` (not equal to)
- `!` (not)
- `&` (and)
- `|` (or)

    ```{r}
    1:3 != 2
    ```

## if

```{r, eval = FALSE}
if (<condition 1>) {
  # do something
  # the only necessary part of if structures
}
else if (<condition 2>) {
  # do something different
}
else {
  # do something else
}
```

## for
- A `for` loop takes an iteration variable, and assigns it successive values from a sequence or a vector.

    ```{r}
    x <- c(90, 87, 60, 45, 50)
    for (i in 1:5) {
      if (x[i] >= 60) {
        print("Pass")
      }
      else {
        print("Fail")
      }
    }
    ```

## while
- `while` loops begin with testing a condition. If it is true, then execute the body. After the execution, the condition will be tested again.
- `while` loops can possibly cause infinite loops if not written properly.

    ```{r}
    i <- 1
    while (i < 4) {
      i <- i + 1
      print(i)
    }
    ```

## Practice
1. Write a loop to revise the unreasonable waist (> 35 inches) in your practiced data frame from centimeters to inches (1 inch = 2.5 cm).
2. The participant with missing waist information has 27 waist. Update your data frame accordingly. DO NOT do `practice.table$waist[10] <- 27`.
3. Export and overwrite "my-success.csv".

# Advanced: Writing Functions

## The first function

```{r}
add <- function(x, y) {
  x + y
}
add(2, 3)
```

## Practice
Try to create a function, which can subset and print out elements greater than a specified number from a numeric vector.

```{r, eval = FALSE}
greaters <- function(x, n) {
  # Your codes here
}
# For example, greaters(1:5, 3) should return 4 and 5.
```

# Applaud Yourself!!