---
title: "Data Viz with R - Shiny (II)"
author: "Yu-Zhen Tu<br>Data engineer at <a href=\"https://www.health2sync.com/tw/\">Health2Sync</a><br>Community Mentor of JHU \"R programming\" on Coursera"
date: "October 1, 2018"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Basic Concepts

## What is Shiny?

According to [Shiny from RStudio](https://shiny.rstudio.com):

- An R package that builds interactive web apps directly from R
- Uses [Bootstrap](http://getbootstrap.com) style
- Can be presented by:
    - Standalone apps on a webpage
    - [R Markdown](http://rmarkdown.rstudio.com/) documents
    - [Dashboards](http://rstudio.github.io/shinydashboard/)
- Can be extended with:
    - [CSS themes](http://rstudio.github.io/shinythemes/)
    - [htmlwidgets](http://www.htmlwidgets.org/)
    - JavaScript [actions](https://github.com/daattali/shinyjs/blob/master/README.md)
    
## Why Shiny?

- Present results (plots, tables, texts, or even value boxes) interactively without leaving analysis pipelines in R
- Easy to share after being deployed (just an URL will do)
- Supported by many other packages, e.g. `leaflet` and `plotly`

## Built-in demo

```{r, echo = TRUE, eval = FALSE}
library(shiny)

ui <- fluidPage(
   # Application title
   titlePanel("Old Faithful Geyser Data"),
   # Sidebar with a slider input for number of bins 
   sidebarLayout(
      sidebarPanel(
         ###<b>
         sliderInput("bins", "Number of bins:",
                     min = 1, max = 50, value = 30)
         ###</b>
      ),
      # Show a plot of the generated distribution
      mainPanel(
         ###<b>
         plotOutput("distPlot")
         ###</b>
      )
   )
)
```

## Built-in demo (cont.)

```{r, echo = TRUE, eval = FALSE}
server <- function(input, output) {
  # generate bins based on input$bins from ui.R
  # draw the histogram with the specified number of bins
  ###<b>
   output$distPlot <- renderPlot({
      x    <- faithful[, 2]
      bins <- seq(min(x), max(x), length.out = input$bins + 1)
      hist(x, breaks = bins, col = 'darkgray', border = 'white')
   })
  ###</b>
}

# Run the application 
shinyApp(ui = ui, server = server)
```

# More Configurations

## More inputs and outputs

- [Shiny cheatsheet](https://www.rstudio.com/resources/cheatsheets/#shiny)
- Provided by other packages, such as `plotly::renderPlotly` and `plotly::plotlyOutput`

## HTML tags in Shiny

- `h1` through `h6`
- `em`, `code`, `image`, `br`...
- Use `tags` list to pick required tag
- Use `HTML` to write HTML codes directly
- BTW, `helpText` to create help texts

## Submit button

- Delayed execution
- Usually used when the app takes much time or resource for calculation
- `submitButton`, or more versatile, `actionButton`

## Reactivity

- Reactive expressions
- Subject to changes in UI inputs
- Wrapped by `reactive` function

    ```{r, echo = TRUE, eval = FALSE}
    # It's like creating a function
    cal_diff <- reactive({
      input$total1 - input$total2
    })

    cal_diff()
    ```

## Combining `ggplot2` into `Shiny`

- Try to create [this app](https://corytu.shinyapps.io/airquality/)
- [airquality_app.R](https://github.com/corytu/RLanguagePresentations/blob/master/docs/airquality_app.R)
- Using characters in `ggplot2` functions
    - `aes_string` in `ggplot`
    - `paste` and `as.formula` in `facet_grid`

# App Deployment

## [shinyapps.io](https://www.shinyapps.io)

- Pros:
    - Can be totally free
    - Simple
    - Scalable (?)
- Cons:
    - Less flexible
    - Limited resources for free plan
- [User Guide for deployment](http://docs.rstudio.com/shinyapps.io/getting-started.html#deploying-applications)

## [Shiny Server](https://www.rstudio.com/products/shiny/shiny-server/) on VM

- [Step-by-step documentation](https://github.com/corytu/OldPopulationProportions/blob/master/setup/howto_deploy_onGCP.md) on Google Cloud Platform

# Cases and Resources

## Case sharing

- [臺銀外匯資料](https://corytu.shinyapps.io/currency-dashboard/)
- [臺灣各鄉鎮市區老化情形](https://corytu.shinyapps.io/old-population-proportions/)

## MOOC and other demos

- [Developing Data Products](https://www.coursera.org/learn/data-products) on Coursera, offered by JHU
- [Shiny Gallery](https://shiny.rstudio.com/gallery/)